Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-02-11 paper.tex
filized methods discussion and other parts

\begin{abstract}
  Wow, we can multiply arrays and tell you how cool it is.
\end{abstract}

\keywords{Arrays, multiplication, addition}

@@

\section{Introduction}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation -- this is startling to observe holy wars on the question which
metric is superior to flare up occasionally,
e.g.,\cite{vogel2022holistic,yudina2023dual}. In fact, we need a single
universal descriptor defined by only one characteristic -- information
content. The latter is an ability of the metric to describe the structure at
hand with 100$\%$ meaning the ability to fully recover the structure back from
this metric. Thus, universal descriptor has to possess following key properties:
(1) high information content, (2) easy to assess, (3) high compression (i.e.,
descriptor is much "smaller" than the structure itself).

The large portion of the research papers mentioned above
\cite{rozenbaum2014,karsanina2015,ledesma2018,derossi2019,portillo2018developing,
  takada2003three,hopkins2013stars,jiao2013,Torquato_book,roding2020predicting,
  KarsaninaEJSS,capek2009,gerke2019tensor,REVpaper,LavrukhinPRE,Adler_recon,Y-T,
  tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,jiao2007,SciRep1,Havelka,Geoderma2018}
is based on so-called correlation functions (CFs). These functions define a
probability that some events are occurring withing the structure (with some
special cases of surface functions), for example, 2-point probability $S_2$
measures the probability that both ends of the line segment are within the same
phase, 2-point cluster function $C_2$ -- that both ends of the line segment are
within the same interconnected cluster of same phase, and lineal path $L_2$
measures the probability that the whole segment lies within the phase (see
\highlight{Fig.1}). Compared to other descriptors, CFs possess a number of
features making them, in our opinion, a perfect candidate for universal
descriptor:
\begin{enumerate}
  \item They contain the majority of other metrics inside. For example, porosity
    is $S_2(0)$ (and at zero correlation length of the majority of correlation
    functions except for surface ones), surface area is represented by in the
    derivative at $S_2(0)$ \cite{debye1957scattering} or the value of
    $F_{ss}(0)$ (or $F_{sss}(0)$). Connectivity is described by cluster
    functions $C_2$ or $C_3$. Pore-size distribution is related to pore-size
    function and partially by $L_2$ and $S_3$. Surface-void functions are
    interconnected with mean curvature and Euler characteristic
    \cite{ma2020generation}. All together, the classic set of correlation
    functions describes both the geometry and topology of the structure at hand;
  \item While the extent to which such classical set describes any arbitrary
    structure is not yet known, compared to other complex descriptors where are
    known pathways to assess the information content of correlation functions --
    there is a methodology for $S_2$ \cite{Gommes2} and approaches for other
    functions \cite{Degeneraty.045306,cherkasov2023towards};
  \item In addition to a possibility to compute correlation functions using
    images or other structural data, i.e., similar to all other descriptors,
    some CFs can be also determined experimentally
    \cite{debye1957scattering,barrall1992nmr,dietrich1995scattering,li2018accurate};
  \item Correlation functions can be scaled \cite{karsaninaPRL} or manipulated,
    for example, to add functions into a single one \cite{moctezuma2002};
  \item As opposed to scalar metrics such as porosity, surface area or Minkowski
    functionals, CFs can be computed in directions and describe anisotropic
    structures;
  \item Numerous rigorous bounds to obtain thermal, elastic, electromagnetic and
    transport properties are known, among which the most known (and least useful
    due to the impractically huge gap between the bounds) are Hashin-Strikman
    bounds \cite{hashin1963variational}. Other expressions based on integration
    of higher order $S_n$ are known and were found to be close to direct
    simulations
    \cite{brown1955solid,beran1965use,milton1981bounds,hlushkou2015effective};
  \item One can control both the size of the CFs-based descriptor and the
    information content via correlation length, number of functions in the set
    (including directions \cite{EPL2}), and by the order of functions (e.g., 2-
    and 3-point versions, or even more points).
\end{enumerate}
This list highlights that correlations functions are indeed possess necessary
properties so satisfy key criteria stated above.

However, while 2-point CFs are computed and used in multitude of studies, higher
order functions, for example, 3-point correlation functions are not that widely
used. The reason for this is straightforward -- 3-point correlation functions
are more involved and, thus, require much more computational resources to
assess. It has been shown that the amount of additional information content with
each increase in number of points $n$ decreases \cite{yao1993high,Gommes2}. The
trade-off between additional information and computational burden is still not
clear. Nonetheless, the computation of 3-point statistics is known to be
advantageous at least in the number of applications as related to cosmology,
astrophysics and turbulence patterns analysis as they can probe non-Gaussian
signatures in the distribution of matter and energy, which are not captured in
2-point counterparts
\cite{TakadaJain,hopkins2013stars,gorbunova2016precessing,yoo2022non}.

@@

The paper is organized as follows: in \cref{sec:methods} we provide
mathematical background and describe methods for computation of 3-point
correlation functions. Verification of these methods along with some results for
real-world samples (sandstone, soil etc.) is described in
\cref{sec:results}. Section \ref{sec:discussion} provides the discussion, a
comparison with other existing methods and summarizes possible directions for
improvement.

@@

Introduce a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function. With this function we can define three additional
correlation functions, which are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as:
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_v(x + x_1) I_v(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_v(x + x_1) I_v(x + x_2) \rangle
\end{align}
Here $I_v$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g. $\mu m^{-3}$, $\mu m^{-2}$ and
$\mu m^{-1}$). These functions describe the interface of the subset $A_n$ and
its spatial configuration with respect to $A_v$.

\subsection{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. Suppose that a finite
countable set $A$ is defined as a multidimensional array. A point $x$ belongs to
a subset $A_n$ if and only if $A[x] = n$. The simplest algorithm below
calculates $S_3$ pointwise in spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \Comment Apply $I_n$ to the input array elementwise.
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and a normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator:
\begin{equation}
  A[x] \leftarrow A[x+z \mod d]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.4\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.4\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

The second approach is to compute the whole correlation map in Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is:
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau.
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500 \times 500$.
Then, considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map is needed even if only one
element of the map in spatial domain is required. The aforementioned limitations
render this approach impractical.

Our implementation fixes $x_1$ and $x_2$ to be parallel to one of the axes $X$,
$Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern}). Thus, the
output has dimensions $D_1 \times D_2$ where $D_1$ and $D_2$ are dimensions of
the input array along selected axes.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\linewidth]{images/pattern.png}
  \caption[]{The pattern used for calculation of $S_3$ function in our
    implementation. Vectors $x_1$ and $x_2$ are parallel to one of the axes $X$,
    $Y$ or $Z$ and have lengths $r_1$ and $r_2$ respectively. A red triangle
    contributes to $S_3(2, 6)$ and blue triangles contribute to $S_3(1, 1)$.}
  \label{fig:pattern}
\end{figure}

Algorithms for three-point surface functions are the same as for $S_3$,
containing only small modifications. The main difference is that we need to
replace an indicator function $I_n$ with an edge detection operator $E_n$
(see \ref{sec:filter}). Then the algorithms are straightforward:
\begin{algorithmic}[1]
  \Procedure{$F_{sss}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

@@

\subsection{Analysis of anisotropic media}
When an analized sample is anisotropic, i.e. it is stretched or oriented along
some specific direction, it may be desired to align one of the catheti of the
scanning pattern \cref{fig:pattern} with that direction. In this case the
anisotropy present in the sample will be featured in a produced correlation
function \textcolor{red}{most well. voobshe eto po-angliyski?}

>>>>

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic reconstruction from limited data.
The classical descriptors are 2-point correlation functions (CFs), but the computation of 3-point statistics is known to be
advantageous in some cases as they can probe non-Gaussian signatures, not captured by their 2-point counterparts.
Moreover, 3-point and CFs with $n>3$ are believed to possess larger information content and provide 
more information about studies structures. In this paper, we have developed algorithms and code to compute 
$S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle triangle pattern. This pattern is much
easier to vectorize and parallelize compared to previously used random triangle or spider-web sampling patterns. Moreover,
it allows to easily implement computations of directional 3-point CFs. The execution times
for the same number of sample are orders of magnitude lower than existing published counterparts. We showed that the volume of
data produced get unwieldy very easily, especially if computations are performed in frequency domain. For these reasons until
information content of different sets of correlation functions with different "n-pointness" is not known, 
advantages of CFs with $n>3$ are not clear. Nonetheless, developed algorithms and code are universal enough to be
easily extendable to any $n$ (with increasing computational and RAM burden). All results are available as part of open source
package \code{CorrelationFunctions.jl} [Postnicov et al., 2024, Comp.Phys.Commun(accepted)]. As described in this paper, 
3-point CFs computations can be immediately applied in a great number of
research applications, for example: 1) flow and transport velocity fields analysis or any data with non-Gaussian signatures,
2) deep learning for structural and physical properties, 3) structure taxonomy and categorization. 
In all these and numerous other potential cases the ability
to compute directional 3-point functions may be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images (while binary structures were used in this paper
for simplicity of explanations).
\end{abstract}

%\keywords{Arrays, multiplication, addition}

@@

\section{Introduction}
\label{sec:intro}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation. In fact, we need a single universal descriptor defined by only one characteristic -- information
content. The latter is an ability of the metric to describe the structure at
hand with 100$\%$ meaning the ability to fully recover the structure back from
this metric. Thus, universal descriptor has to possess following key properties:
(1) high information content, (2) ease of assess, (3) high compression ratio (i.e.,
descriptor is much "smaller" than the structure itself).

The large portion of the research papers mentioned above
\cite{rozenbaum2014,karsanina2015,ledesma2018,derossi2019,portillo2018developing,
  takada2003three,hopkins2013stars,jiao2013,Torquato_book,roding2020predicting,
  KarsaninaEJSS,capek2009,gerke2019tensor,REVpaper,LavrukhinPRE,Adler_recon,Y-T,
  tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,jiao2007,SciRep1,Havelka,Geoderma2018}
is based on so-called correlation functions (CFs). These functions define a
probability that some events are occurring withing the structure (with some
special cases of surface functions), for example, 2-point probability $S_2$
measures the probability that both ends of the line segment are within the same
phase, 2-point cluster function $C_2$ -- that both ends of the line segment are
within the same interconnected cluster of same phase, and lineal path $L_2$
measures the probability that the whole segment lies within the phase (see
\highlight{Fig.1}). Compared to other descriptors, CFs possess a number of
features making them, in our opinion, a perfect candidate for universal
descriptor:
\begin{enumerate}
  \item They contain the majority of other metrics inside. For example, porosity
    is $S_2(0)$ (and at zero correlation length of the majority of correlation
    functions except for surface ones), surface area is represented by in the
    derivative at $S_2(0)$ \cite{debye1957scattering}. Connectivity is described by cluster
    functions $C_2$ or $C_3$. Pore-size distribution is related to pore-size
    function and partially by $L_2$ and $S_3$. Surface-void functions are
    interconnected with mean curvature and Euler characteristic
    \cite{ma2020generation}. Altogether, the classic set of correlation
    functions describes both the geometry and topology of the structure at hand;
  \item While the extent to which such classical set describes any arbitrary
    structure is not yet known, compared to other complex descriptors where are
    known pathways to assess the information content of correlation functions --
    there is a methodology for $S_2$ \cite{Gommes2} and approaches for other
    functions \cite{Degeneraty.045306,cherkasov2023towards};
  \item In addition to a possibility to compute correlation functions using
    images or other structural data, i.e., similar to all other descriptors,
    some CFs can be also determined experimentally
    \cite{debye1957scattering,barrall1992nmr,dietrich1995scattering,li2018accurate};
  \item Correlation functions can be scaled \cite{karsaninaPRL} or manipulated,
    for example, to add functions into a single one \cite{moctezuma2002};
  \item As opposed to scalar metrics such as porosity, surface area or Minkowski
    functionals, CFs can be computed in directions and describe anisotropic
    structures;
  \item Numerous rigorous bounds to obtain thermal, elastic, electromagnetic and
    transport properties are known, among which the most known (and least useful
    due to the impractically huge gap between the bounds) are Hashin-Strikman
    bounds \cite{hashin1963variational}. Other expressions based on integration
    of higher order $S_n$ are known and were found to be close to direct
    simulations
    \cite{brown1955solid,beran1965use,milton1981bounds,hlushkou2015effective};
  \item One can control both the size of the CFs-based descriptor and the
    information content via correlation length, number of functions in the set
    (including directions \cite{EPL2}), and by the order of functions (e.g., 2-
    and 3-point versions, or even more points).
\end{enumerate}
This list highlights that correlations functions are indeed possess necessary
properties so satisfy key criteria stated above.

However, while 2-point CFs are computed and used in multitude of studies, higher
order functions, for example, 3-point correlation functions are not that widely
used. The reason for this is straightforward -- 3-point correlation functions
are more involved and, thus, require much more computational resources to
assess. It has been shown that the amount of additional information content with
each increase in number of points $n$ decreases \cite{yao1993high,Gommes2}. The
trade-off between additional information and computational burden is still not
clear. Nonetheless, the computation of 3-point statistics is known to be
advantageous at least in the number of applications as related to cosmology,
astrophysics and turbulence patterns analysis as they can probe non-Gaussian
signatures in the distribution of matter and energy, which are not captured by their
2-point counterparts
\cite{TakadaJain,hopkins2013stars,gorbunova2016precessing,yoo2022non}.

@@

The paper is organized as follows: in \cref{sec:methods} we provide
mathematical background and describe methods for computation of 3-point
correlation functions. Verification of these methods along with some results for
real-world samples (sandstone, soil, etc.) is described in
\cref{sec:results}. Section \ref{sec:discussion} provides the discussion, a
comparison with other existing methods and summarizes possible directions for
future research.

@@

By introducing a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function, we can define three additional
correlation functions -- these are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as:
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_v(x + x_1) I_v(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_v(x + x_1) I_v(x + x_2) \rangle
\end{align}
Here $I_v$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g., can be measured in $\mu m^{-3}$, $\mu m^{-2}$ and
$\mu m^{-1}$ or other convenient dimensionless units depending on the problem to be solved). 
These functions describe the interface of the subset $A_n$ and
its spatial configuration with respect to $A_v$.

\subsection{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. Suppose that a finite
countable set $A$ is defined as a multidimensional array. A point $x$ belongs to
a subset $A_n$ if and only if $A[x] = n$. The simplest algorithm below
calculates $S_3$ pointwise in the spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \Comment Apply $I_n$ to the input array elementwise.
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and a normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator:
\begin{equation}
  A[x] \leftarrow A[x+z \mod d]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in Appendix
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.4\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.4\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

The second approach is to compute the whole correlation map in the Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is:
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau.
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500 \times 500$.
Then, considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map has to be computed even if only one
element of the map in spatial domain is required. The aforementioned limitations
render this approach impractical.

Our implementation fixes $x_1$ and $x_2$ to be parallel to one of the axes $X$,
$Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern}) \cite{dimitrakopoulos2010high}. Thus, the
output has dimensions $D_1 \times D_2$ where $D_1$ and $D_2$ are dimensions of
the input array along selected axes. Note that shift operators with such $x_1$ and $x_2$
represent a 3-point function computed within a 3D domain, but along a predefined direction (see below).
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\linewidth]{images/pattern.png}
  \caption[]{The pattern used for calculation of $S_3$ function in our
    implementation. Vectors $x_1$ and $x_2$ are parallel to one of the axes $X$,
    $Y$ or $Z$ and have lengths $r_1$ and $r_2$ respectively. A red triangle
    contributes to $S_3(2, 6)$ and blue triangles contribute to $S_3(1, 1)$.}
  \label{fig:pattern}
\end{figure}

Algorithms for three-point surface functions are the same as for $S_3$,
containing only small modifications. The main difference is that we need to
replace an indicator function $I_n$ with an edge detection operator $E_n$
(see Appendix \ref{sec:filter}). Then the algorithms are straightforward:
\begin{algorithmic}[1]
  \Procedure{$F_{sss}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

@@

\subsection{Analysis of anisotropic media}
When an analyzed sample is anisotropic, i.e., it is stretched or oriented along
some specific direction, it may be beneficial to align one of the cathetus of 
sampling triangle \cref{fig:pattern} with that direction. In this case the
anisotropy present in the sample will be properly captured by computed 3-pint correlation function.

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}.
\begin{figure*}[tp]
  \centering
  \subfigure[An original image of an anisotropic pore.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso.png}
    \label{fig:aniso-unrot}}
  \hfill
  \subfigure[A rotated image of the pore, so it is elongated along the axis
    $X$.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso-rot.png}
    \label{fig:aniso-rot}}
  \vskip\baselineskip
  \subfigure[$S_3$ function for the original image]{
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}}
  \hfill
  \subfigure[$S_3$ function for the rotated image. A vector $(0.92, -0.4)$ is a
    direction of elongation of the pore.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}
    \label{fig:aniso-rot-s3}}
  \hfill
  \subfigure[$S_3$ function averaged across 100 random rotations.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
    \label{fig:aniso-avg}}
  \caption[]{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. A result on \cref{fig:aniso-avg} averaged
    across rotations is similar to the algorithm described in \cite{berryman1988,SMITH1988176,malmir2018}. \highlight{Make empty image over (e)}}
  \label{fig:aniso}
\end{figure*}

\subsection{Directional and averaged computations}
When a sample is known to be isotropic, the correlation function may be
calculated multiple times, each time applying a random rotation of the sample,
and then averaged across rotations to gather more statistics. This process is
illustrated on \cref{fig:workflow}. This way we obtain similar results to sampling procedures
used in \cite{berryman1988,SMITH1988176,malmir2018}.

Another approach is to fix a number of directions, for example, orthogonal and
diagonal ones \cite{EPL1,EPL2,CFsjlpaper}. In other words, the sample is rotated with periodic boundary conditions
to orient along predefined directions. After each rotation we compute correlation functions
as described above and obtain a number of matrices $D_1 \times D_2$ that are then
stacked to represent 3-point correlation functions computed in directions \cref{fig:workflow}.

To summarize, \code{CorrelationFunctions.jl} allows to compute 3-point correlation functions
averaged in all directions (by randomly rotating the sample), along predefined number of directions
(e.g., orthogonal and diagonal by rotation in these directions)
or along the direction of major anisotropy (that can be determined using \code{$detect\_anisotropy$} function).
Full correlation map for 3-point functions is impractical due its huge size and is not implemented.

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.6\linewidth]{images/workflow.png}
  \caption[]{A possible workflow for computation of correlation functions
    (clockwise from top left to bottom right ): 1.~rotating a sample in random
    directions (which is equivalent to rotating the sampling pattern),
    2.~calculating the correlation function for each direction, 3.~avegaring
    results across rotations, 4.~representing the result as a family of 2D
    plots. \highlight{Show "random rotation" at the slice, "Stack of directional functions" on the right-upper.}}
  \label{fig:workflow}
\end{figure*}
